Convergent Evolution During Local Adaptation to Patchy Landscapes

### 2015-08-29 patchy-review-responses-second-round.tex

\vskip 2em

{\bf Peter Ralph, and\\
Graham Coop.}\\
\end{minipage}

\vskip 4em

The main change,
which addresses most of the concerns,
is that we uncovered an error in the derivation of equation \eqref{eqn:migrate},
that supplied the missing factor of area noted by the reviewers.
The error occurred in the transition between section \nameref{ss:heuristics} 
and section \nameref{ss:hitting_occupation}
because in the former we had tried to keep things simple
by not including shape of the patch in
gestalt equations \eqref{eqn:gestalt_q} and \eqref{eqn:gestalt_migrate},
but then (more correctly) treated the shape in the latter;
effectively we were treating the patch as very small in one place
and large in another.
We've rewritten the argument in \nameref{ss:heuristics} 
to deal explicitly with the integral over the whole patch,
which also necessitated adding a new (short) appendix \nameref{apx:qS}.
Section \nameref{ss:hitting_occupation} required very little change.

@@

The main change,
which addresses most of the concerns,
is that we uncovered an error in the derivation of equation \eqref{eqn:migrate},
that supplied the missing factor of area noted by the reviewers.
The error occurred in the transition between section \nameref{ss:heuristics} 
and section \nameref{ss:hitting_occupation}
because in the former we had tried to keep things simple
by not including shape of the patch in
gestalt equations \eqref{eqn:gestalt_q} and \eqref{eqn:gestalt_migrate},
but then (more correctly) treated the shape in the latter;
effectively we were treating the patch as very small in one place
and large in another.
We've rewritten the argument in \nameref{ss:heuristics} 
to deal explicitly with the integral over the whole patch,
which also necessitated adding a new (short) appendix \nameref{apx:qS}.
Section \nameref{ss:hitting_occupation} required very little change.

%%%%%%%%%%%%%%
\reviewersection

%%%%%%
\point{}{
    The issue of dimensions and geometry of the target patch is the main difficulty I have with the paper. I just don't see how your equation \eqref{eqn:migrate} can be correct. It has dimensions of 1/time/area when it should be 1/time. Somehow the target area needs to play a role. Reviewer 2 has raised similar issues.
}

\reply
Now that we've found the problem in the derivation,
we heartily agree;
the missing factor was, indeed, a multiplicative factor of area.
(See above for more on this.)

%%%%%%
\point{}{
Regarding the simulations presented in Fig 4, I don't think they strongly support the functional form of $\migrate$. Points with very similar observed migration time have predicted migration times that differ by a factor of a 100 (the more extreme are outside the validity of the approximations, but still I don't think this figure is strong evidence that the math is correct.)
}

\reply
We think the main issue here was inclusion of points outside the validity of the approximations,
since of the parameters falling in the range of validity discussed in \nameref{ss:assumptions},
the largest discrepancy was a factor of 7, not 100.
We've re-made this figure only using parameter combinations that fall within the range of validity
($R > \sigma/\sqrt{s_m}$, patch width larger than $\sigma/\sqrt{s_m} \atan(\sqrt{s_m/s_b})$, and $2 s_m \sigma N > 1$);
%         ( R*sqrt(abs(sm))/sigma > 1 ) & 
%         ( size2 > (sigma/sqrt(abs(sm)))*atan(sqrt(abs(sm/gb))) ) &
%         ( abs(sm) * 2 * sigma * N > 1 )
the figures that include all simulation results are still included in the supplement
(and hopefully will inspire theory that is applicable in the other regions).

@@

%%%%%%
\point{eq.~56 is wrong}{
    as easily seen with $\Theta=0$ \ldots
    so eq.~56 has an extra 2, resulting from a wrong expansion of $\sqrt{1+x}$,
    and so eq.~57--61 are also wrong.
}

\reply
Thanks for catching that mistake.  We removed the extra factor of 2 (section~\nameref{apx:outflux}).

%%%%%%
\point{time in transit}{
    One can obtain these results by a simple saddle point approximation:
    \begin{align*}
        \P\{ \text{getting to $x$ in time $t$} \} &= \P\{\text{diffusion to $x$ in time $t$}\} \P\{\text{surviving}\} \\
                                                  &= \frac{1}{(2\pi\sigma^2)^{d/2}} e^{-\frac{x^2}{2\sigma^2 t}-st}
    \end{align*}
    the exponent is minimal when
    \[
        \frac{x^2}{2\sigma^2 t^2} = s \quad \Rightarrow \quad t^* = \frac{x}{\sigma\sqrt{2s}},
    \]
    this gives your eq.~17.
    The variance is simply the inverse of the second derivative of the exponent
    \[
        \frac{1}{\var[\tau]} = \frac{x^2}{\sigma^2 (t^*)^3} = \frac{\sigma}{x(2s)^{3/2}} .
    \]
}

\reply
This is a nice point;
perhaps which approach should be included
comes down to personal taste
as to whether ``saddle point approximation'' or ``Laplace transform'' is a simpler idea?  
We've left this as-is.

### 2015-08-29 patchy-selection-paper.tex
edits

\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\Exp}{Exp}
\DeclareMathOperator{\sn}{sn}
\DeclareMathOperator{\erf}{erf}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\one}{\mathbf{1}}
\newcommand{\calE}{\mathcal{E}}
\newcommand{\calC}{\mathcal{C}}
\newcommand{\dconv}{\xrightarrow{d}}
\newcommand{\deq}{\stackrel{\scriptscriptstyle{d}}{=}}
\newcommand{\given}{\,\vert\,}

@@

Suppose the family arrives with $K$ individuals.
The probability that at least one individual leaves a lineage that establishes in the patch
is $1-(1-p_e)^K$.
On the other hand, in the absence of the patch of new habitat,
each of the $K$ newly arrived individuals would leave behind a lineage with 
mean total size $\sum_{t=0}^\infty e^{-t s_m} \approx 1/s_m$,
roughly half of which would stay in the new patch.
Therefore, the expected contribution of the family to the occupation density is $K/2s_m$.
(Readers who feel uncomfortable about the geometry of the patch,
or the distinction between occupation density and number of individuals
are encouraged to take a detour through section \secref{ss:hitting_occupation}, below.)

@@

\begin{figure}[ht!]
  \begin{center}
      % made in analyze-sims.R
      \includegraphics{times-predicted-observed}
  \end{center}
  \caption{
      \textbf{(left)} 
      % XXX HERE
      Time until adaptation by mutation in an isolated patch for different parameter combinations,
      as observed in simulations (vertical axis),
      and as predicted from equation~\eqref{eqn:mutrate} (horizontal axis).
      \textbf{(right)} 
      Time until adaptation by migration from an already occupied patch to a new patch,
      as observed in simulations (vertical axis),
      and as predicted from equation~\eqref{eqn:migrate}, with $C=5$ (horizontal axis).
      For each,
      parameter combinations and numbers of simulations are in supplemental tables \ref{stab:mutation_params} and \ref{stab:migration_params};
      shown are the median (points) and interquartile range (lines)
      of the time until the patch had at least 100 adapted individuals,
      for each unique parameter combination.
  }   \label{fig:sim_times}
\end{figure}

@@

First consider equation \eqref{eqn:gestalt_q} for the equilibrium frequency.
Suppose that $Z$ is one such spatial branching process as above in section \secref{ss:migrant_math}, 
started at time 0 with a single individual at $x$,
and write $S$ for the new patch, which has area $A$ and is centered at $y$.
The \emph{mean occupation measure} of $Z$ in the region $S$
which we denote $u(x,S)$,
% and denote by $Z_t(A)$ is the number of descendant individuals alive at time $t$ living in the spatial region $A$.
% The \emph{mean occupation density} of $Z$ at location $y$
can be thought of informally \linelabel{rr:density}
as the expected total number of offspring of a family beginning at $x$ that ever live in $S$.
% It is defined to be the density of $\E[\int_0^t Z_t(\cdot) dt]$,
% i.e.\ the function $u(x,y)$ that, integrated over a set of locations $S$, 
% gives the expected total number of individuals ever living in $S$:
% \begin{align} \label{eqn:occupation_density_defn}
%     \E\left[ \int_0^\infty Z_t(S) dt \right] = \int_S u(x,y) dy .
% \end{align}
% % Said another way, $u(x,y)$ is the expected number of offspring of a progenitor at $x$ that will ever be observed near $y$.
% Since reproduction and spatial motion are independent,
% $u(x,y)$ is the sum across all generations of the expected number of individuals alive in that generation,
% multiplied by the probability density that spatial motion takes a lineage from $x$ to $y$,
% i.e.\
% \begin{align}
%     u(x,y) = \int_0^\infty e^{-s_m t} p_t(x,y) dt .
% \end{align}
Let $q(S) = \int_S q(y) dy$ denote the total equilibrium frequency in $S$.
This is decomposed in expression \eqref{eqn:gestalt_q} 
as the sum of mean occupation densities of a constant outflux of branching processes
from $\partial B(r_0)$:
\begin{align} \label{eqn:occupation_integral}
    \rho q(S) &= \int_{\partial B(r_0)} \gamma(x) u(x,S) dx  . 
\end{align}

Now we will decompose $u(x,S)$ under the assumption that 
the marginal distribution of the spatial motion of a single lineage is Brownian.
Let $B_t$ be a Brownian motion with variance $\sigma^2$, 
and $\tau_\dagger$ an independent Exponential($s_m$) time.
The mean occupation time of $Z$ spent in a region $S$ is,
\begin{align}
    u(x,S) &= \E\left[\int_0^\infty Z_t(S) dt \right] \\
           &= \int_0^\infty \E[Z_t] p_t(x,S) dt \\
        &= \int_0^\infty e^{-s_m t} p_t(x,S) dt \\
        &= \int_0^\infty \P^x\{ \tau_\dagger > t \; \& \; B_t \in S \} dt \\
        &= \E^x\left[ \int_0^{\tau_\dagger} \one_S(B_t) dt \right]  ,
\end{align}
where $\P^x$ gives probabilities for Brownian motion begun at $x$ (i.e.\ $B_0=x$),
and likewise $\E^x$.
If we define
$\tau_S$ to be the hitting time of $S$ by the Brownian motion $B$,
and $\mu_S(x)$ to be the hitting distribution of $\partial S$ by $B_{\tau_S}$ conditioned on $\tau_S < \tau_\dagger$,
by the strong Markov property this is equal to
\begin{align}
    u(x,S) % &= \E^x[ \int_0^{\tau_\dagger} \one_S(B_t) dt ] \\
           &= \P^x\{ \tau_S < \tau_\dagger \}  \E^{\mu_S(x)}\left[ \int_0^{\tau_\dagger} \one_S(B_t) dt \right] \\
           &= \P^x\{ \tau_S < \tau_\dagger \}  g(A)  ,
\end{align}
where now $\E^\mu$ denotes expectations for Brownian motion for which the distribution of $B_0$ is $\mu$, 
% provided $S$ is not pathological.
and $g(A)$ is defined to be the latter expectation,
which does not depend on $x$ since $S$ is circular with area $A$.

@@

XXX CHECK FACTORS OF $\rho$ BELOW XXX

@@

% Intuitively, $h_A(x,y)$ counts each family once if it hits the patch,
% while the occupation density $u(x,y)$ counts the total amount of time that the family spends in the patch.
% Since the expected total time spent in the patch does not depend on the distance at which the family started
% if we know that it survived to hit the patch,
% it is intuitively reasonable that $u(x,y)$ and $h_A(x,y)$ are simply related.
% Following this intuition,
% and using $(1-k_e(t)) \approx e^{-s_m t}/\E[K]$,
% we show below that
% \begin{align} 
%     h_A(x,y) \approx \frac{ u(x,y) }{ \E[K] g(A) } ,
% \end{align}
% where $g(A)$ is the total expected occupation time of the patch given the family arrives.
% Recall that $\E[K]$ is the expected family size after a large number of generations, given survival.
Since the integrands in expressions \eqref{eqn:occupation_integral} and \eqref{eqn:migrate_integral} only differ by a factor
that (at least asymptotically) does not depend on the distance between $x$ and $y$,
we can obtain the migration rate by multiplying the equilibrium frequency by this factor.
XXX NOT RIGHT XXX
The result will not depend on $A$, 
because although $f(A)$ and $g(A)$ in principle depend on the patch size (and geometry),
the dependence is very weak.
For instance, the width of a circular patch only very weakly affects the chance 
of establishment of a new migrant that appears on its edge,
as long as the patch is large enough.

@@

\section*{Acknowledgements}
We would like to heartily thank Joachim Hermisson for a number of insightful suggestions,
and catching the algebra error present in the first version of this paper on the bioRxiv,
as well as the two anonymous reviewers.
We would also like to acknowledge Michael Nachman, Steve Evans, and Yaniv Brandvain for useful discussions, 
as well as Gideon Bradburd, and Simon Aeschbacher and the rest of the Coop lab
for helpful comments and discussion of the manuscript.
This work was supported by grants from the
National Science Foundation under Grant No. 1262645 to PLR and GC and by the
National Institute of General Medical Sciences of the National
Institutes of Health under award numbers NIH RO1GM83098 and
RO1GM107374 to GC.

>>>>

\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\Exp}{Exp}
\DeclareMathOperator{\sn}{sn}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\atan}{atan}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\one}{\mathbf{1}}
\newcommand{\calE}{\mathcal{E}}
\newcommand{\calC}{\mathcal{C}}
\newcommand{\dconv}{\xrightarrow{d}}
\newcommand{\deq}{\stackrel{\scriptscriptstyle{d}}{=}}
\newcommand{\given}{\,\vert\,}

@@

Suppose the family arrives with $K$ individuals.
The probability that at least one individual leaves a lineage that establishes in the patch
is $1-(1-p_e)^K$.
On the other hand, in the absence of the patch of new habitat,
each of the $K$ newly arrived individuals would leave behind a lineage with 
mean total size $\sum_{t=0}^\infty e^{-t s_m} \approx 1/s_m$,
roughly half of which would stay in the new patch.
Therefore, the expected contribution of the family to the occupation density is $K/2s_m$.
(For more careful consideration of the geometry of the patch
and the distinction between occupation density and number of individuals,
see section \secref{ss:hitting_occupation}, below.)

@@

\begin{figure}[ht!]
  \begin{center}
      % made in analyze-sims.R
      \includegraphics{times-predicted-observed}
  \end{center}
  \caption{
      \textbf{(left)} 
      % XXX HERE
      Time until adaptation by mutation in an isolated patch for different parameter combinations,
      as observed in simulations (vertical axis),
      and as predicted from equation~\eqref{eqn:mutrate} (horizontal axis).
      \textbf{(right)} 
      Time until adaptation by migration from an already occupied patch to a new patch,
      as observed in simulations (vertical axis),
      and as predicted from equation~\eqref{eqn:migrate}, with $C=5$ (horizontal axis).
      For each,
      parameter combinations and numbers of simulations are in supplemental tables \ref{stab:mutation_params} and \ref{stab:migration_params};
      shown are the median (points) and interquartile range (lines)
      of the time until the patch had at least 100 adapted individuals,
      for each unique parameter combination.
      Only parameter combinations to which the theory here is applicable were used
      (precisely, if
      $R > \sigma/\sqrt{s_m}$, 
      patch width larger than $\sigma/\sqrt{s_m} \atan(\sqrt{s_m/s_p})$, 
      and $s_m > 2 \sigma \rho$);
      the remaining combinations are shown in figures
      \ref{sfig:sim_migration_times} and \ref{sfig:sim_mutation_times}.
  }   \label{fig:sim_times}
\end{figure}

@@

First consider equation \eqref{eqn:gestalt_q} for the equilibrium frequency.
Suppose that $Z$ is one such spatial branching process as above in section \secref{ss:migrant_math}, 
started at time 0 with a single individual at $x$,
and write $S$ for the new patch, which has area $A$ and is centered at $y$.
The \emph{mean occupation measure} of $Z$ in the region $S$,
which we denote $u(x,S)$,
% and denote by $Z_t(A)$ is the number of descendant individuals alive at time $t$ living in the spatial region $A$.
% The \emph{mean occupation density} of $Z$ at location $y$
can be thought of informally \linelabel{rr:density}
as the expected total number of offspring of a family beginning at $x$ that ever live in $S$.
% It is defined to be the density of $\E[\int_0^t Z_t(\cdot) dt]$,
% i.e.\ the function $u(x,y)$ that, integrated over a set of locations $S$, 
% gives the expected total number of individuals ever living in $S$:
% \begin{align} \label{eqn:occupation_density_defn}
%     \E\left[ \int_0^\infty Z_t(S) dt \right] = \int_S u(x,y) dy .
% \end{align}
% % Said another way, $u(x,y)$ is the expected number of offspring of a progenitor at $x$ that will ever be observed near $y$.
% Since reproduction and spatial motion are independent,
% $u(x,y)$ is the sum across all generations of the expected number of individuals alive in that generation,
% multiplied by the probability density that spatial motion takes a lineage from $x$ to $y$,
% i.e.\
% \begin{align}
%     u(x,y) = \int_0^\infty e^{-s_m t} p_t(x,y) dt .
% \end{align}
Let $q(S) = \int_S q(y) dy$ denote the total equilibrium frequency in $S$.
This is decomposed in expression \eqref{eqn:gestalt_q} 
as the sum of mean occupation densities of a constant outflux of branching processes
from $\partial B(r_0)$:
\begin{align} \label{eqn:occupation_integral}
    \rho q(S) &= \int_{\partial B(r_0)} \gamma(x) u(x,S) dx  . 
\end{align}

Now we will decompose $u(x,S)$ under the assumption that 
the marginal distribution of the spatial motion of a single lineage is Brownian.
Let $B_t$ be a Brownian motion with variance $\sigma^2$, 
and $\tau_\dagger$ an independent Exponential($s_m$) time.
The mean occupation time of $Z$ spent in a region $S$ is,
\begin{align}
    u(x,S) &= \E\left[\int_0^\infty Z_t(S) dt \right] \\
           &= \int_0^\infty \E[Z_t] p_t(x,S) dt \\
        &= \int_0^\infty e^{-s_m t} p_t(x,S) dt \\
        &= \int_0^\infty \P^x\{ \tau_\dagger > t \; \& \; B_t \in S \} dt \\
        &= \E^x\left[ \int_0^{\tau_\dagger} \one_S(B_t) dt \right]  ,
\end{align}
where $\P^x$ gives probabilities for Brownian motion begun at $x$ (i.e.\ $B_0=x$),
and likewise $\E^x$.
If we define
$\tau_S$ to be the hitting time of $S$ by the Brownian motion $B$,
and $\mu_S(x)$ to be the hitting distribution of $\partial S$ by $B_{\tau_S}$ conditioned on $\tau_S < \tau_\dagger$,
by the strong Markov property this is equal to
\begin{align}
    u(x,S) % &= \E^x[ \int_0^{\tau_\dagger} \one_S(B_t) dt ] \\
           &= \P^x\{ \tau_S < \tau_\dagger \}  \E^{\mu_S(x)}\left[ \int_0^{\tau_\dagger} \one_S(B_t) dt \right] \\
           &= \P^x\{ \tau_S < \tau_\dagger \}  g(A)  ,
\end{align}
where now $\E^\mu$ denotes expectations for Brownian motion for which the distribution of $B_0$ is $\mu$, 
% provided $S$ is not pathological.
and $g(A)$ is defined to be the latter expectation,
which does not depend on $x$ if $S$ is circular (with area $A$).

@@

@@

% Intuitively, $h_A(x,y)$ counts each family once if it hits the patch,
% while the occupation density $u(x,y)$ counts the total amount of time that the family spends in the patch.
% Since the expected total time spent in the patch does not depend on the distance at which the family started
% if we know that it survived to hit the patch,
% it is intuitively reasonable that $u(x,y)$ and $h_A(x,y)$ are simply related.
% Following this intuition,
% and using $(1-k_e(t)) \approx e^{-s_m t}/\E[K]$,
% we show below that
% \begin{align} 
%     h_A(x,y) \approx \frac{ u(x,y) }{ \E[K] g(A) } ,
% \end{align}
% where $g(A)$ is the total expected occupation time of the patch given the family arrives.
% Recall that $\E[K]$ is the expected family size after a large number of generations, given survival.
Since the integrands in expressions \eqref{eqn:occupation_integral} and \eqref{eqn:migrate_integral} only differ by a factor
that (at least asymptotically) does not depend on the distance between $x$ and $y$,
we can obtain the migration rate by multiplying the equilibrium frequency by this factor.
This factor will not depend on $A$, 
because although $f(A)$ and $g(A)$ in principle depend on the patch size (and geometry),
the dependence is very weak.
For instance, the width of a circular patch only very weakly affects the chance 
of establishment of a new migrant that appears on its edge,
as long as the patch is large enough.

@@

\section*{Acknowledgements}
We would like to heartily thank Joachim Hermisson for a number of insightful suggestions,
and catching the algebra error present in the first version of this paper on the bioRxiv,
as well as the two anonymous reviewers.
We would also like to acknowledge Michael Nachman, Steve Evans, and Yaniv Brandvain for useful discussions, 
as well as Gideon Bradburd, and Simon Aeschbacher and the rest of the Coop lab
for helpful comments and discussion of the manuscript.
This work was supported by grants from the
National Science Foundation under Grant No.\ 1262645 to PLR and GC and by the
National Institute of General Medical Sciences of the National
Institutes of Health under award numbers NIH RO1GM83098 and
RO1GM107374 to GC.
